\bf Visual Diagnostics of Dynamic Social Network Models

### 2019-07-08 paper-child.Rnw
stuff from last week

Dynamic network models do not have as rich of a history as static network models, largely due to the historical lack of data and available analytical tools \citep{goldenberg09}. With the advent of social media and ever-increasing computational power, however, more and more researchers are turning toward dynamic network models.  Some of the classical models, such as the Erd\"{o}s-R\'{e}nyi model and the small-world model of \citet{watts} are pseudo-dynamic because they could be applied in a dynamic way but rarely are \citep{goldenberg09}.  
% talk up dynamic social network models. they are much more realistic. they don't treat nodes as exchangeable, they allow the nodes to affect the ties. They account for both structural and covariate effects. They are very flexible. They are the ideal model because they are the only ones that have any chance of actually capturing the mechanisms. 
Fully dynamic social network models, however, are more complicated: they incorporate both the network structure and its changes in time.  If we can capture this change mechanism in time, we can truly begin to model the underlying network dynamics that drive social connections. A comprehensive review of dynamic network models is presented in \citet{golenberg09}.

\par We are particularly interested one truly dynamic family of network models: continuous time Markov chain (CTMC) models, first introduced in \citet{ctmcorig}. We chose these models because they correspond to our intuition about how social networks are formed, and they are very flexible and widely applicable. In these models, network changes are modeled hierarchically at the node level: nodes are given opportunities to change via a rate function, then the node chooses which edge to change according to its utility function. %first step is the changes happen at the smallest possible level (one edge between two nodes), and subsequent changes condition on past changes. 
In each step of the hierarchy, many possible effects can be implemented to capture the change mechanism. For example,  the rate function could be a simple waiting-time model, or it could depend on a node's indegree or outdegree: we may have reason to believe that the more "popular" nodes change their ties more frequently. In addition, we can incorporate node covariate information into the hierarchy, such as gender or education level of people in a social network. For instance, the genders of two people could effect whether and how quickly they establish a relationship, and people will likely form more relationships with those who share their education level. Thus, the transition probabilities between two states of a network depend on the network's structure, characteristics of the nodes, or both \citep{saompaper}.  For example, \citet{depression} use a CTMC model to learn about the tendency of depressed adolescents to form friendships and found that depressed students tended to avoid creating new friendships over time. The CTMC model family can also be expanded to model node behavior, as in \citet{chyzh2016}, who applies CTMC models to show that more indirect trade relationships in a country decrease its overall respect for human rights. % This means that node-level covariate effects, in addition to commonly used effects such as reciprocity or transitivity, can be incorporated into the model. 
% Cite some of olga's work. what can you do with these models? cite other work with social networks. 
Thus, with CTMC models we can examine assumptions made about social networks, i.e. that people with common qualities are more likely to become friends. %For example, we may assume that teenagers who drink heavily are more likely to be friends with other teenagers who drink heavily.  We can assess our assumptions on shared interests into the modeling process through a particular subfamily of CTMC models known as stochastic actor-oriented models (SAOMs) \citep{saompaper}.

\par With the flexibility of CTMC models comes increased estimation difficulty \citep{saomreview}. Likelihood functions quickly become very complex due to the inherent dependency structure in the data, and therefore computational methods, such as Markov chain Monte Carlo (MCMC), are used to fit models. Methods for maximum likelihood estimation (MLE) for these models were not introduced until \citet{saomsml2010}, and method of moments (Mom) estimation is frequently used \citep{RSiena}. %SAOMs are typically fit to dynamic social network data using a series of MCMC phases for finding method of moments (MoM) estimates of model parameters. 
For CTMC parameter estimation, we use the software SIENA, and its \texttt{R} implementation \texttt{RSiena} \citep{rstats, RSiena}. This software is a considerable contribution to the field of social network analysis, allowing for estimation of structural effects, node covariate effects, as well as modeling both network and node changes. This extensive parameter estimation process is largely hidden from the user, and the primary outputs are parameter estimates. Direct simulation from a given network model with fixed parameter values is also made easy with \texttt{RSiena}. This software opens up a great deal of possibilities for network modeling and simulation.% need to emphasize the need for visualization.

% ok. we know why ctmcs but WHY visualize them? where is the motivation?? who cares and why? 
% talk about lack of goodness of fit diagnostics. 
After choosing the CTMC models for dynamic social networks, we would like to determine how well they fit data. We chose these models because they match our intuition about how social networks form in reality, and after we fit a model, we will want to assess its goodness-of-fit. In statistical network models generally, the lack of large-sample theory leads to a lack of goodness-of-fit tests \citep{goldenberg09}. Most methods for assessing goodness-of-fit for network models, therefore, involve comparing statistics computed on simulated networks to the same statistics computed on the data, for example in \citet{hunter2008}. More recently, \citet{gofschwein} used score-type tests for forward model selection. The \texttt{RSiena} software also contains methods for goodness-of-fit based on network statistics, such as outdegree distribution, where the data values are compared to simulated values as in \citet{hunter2008} \citep{RSiena}. % need to transition to model vis. % say something about how important visualization is in network analysis.

We propose new methods for goodness-of-fit and other diagnostics that use visualization as a supplement to computational methods. Network visualization, also called graph drawing, is a prominent subfield of network analysis with a rich history: early drawings of complete graphs date back to 1280 CE. We use this visualization foundation to combine modeling with visualization. Note that we do not propose new methods of visualizing networks, and for a complete review of the field of graph drawing we direct the reader to \citet{graphdrawing}. What is novel in our work is the use of common visualization methods to diagnose network models. By translating the model visualization work of \citet{modelvis} to the dynamic network model space, we % do something awesome

@@

%\noindent \textbf{Network Data Structure:} 
Network data across fields have similar structure: the data consist of units of observation and connections between those units. In a social network, the units of observations, called \textit{nodes} or \textit{actors}, are people, while the connections, called \textit{edges} or \textit{ties}, are their relationships with each other. %Networks change over time, like when new relationships are formed in a social network. 
Other network data include inherent variables of interest on the nodes, such as age or gender, and the type of relationship (friends, coworkers, etc) that an edge represents. Dynamic networks are observed at many discrete points in time, where each observation of a network is often called a "wave" of data.
%These two network data structures pose unique problems to analysts, such as, ``How does the strength of a tie between two nodes affect the overall structure of the network?" or ``Do differences in node variables affect the formation of edges?" We propose to answer these questions and more through visual exploration of network data and models. \\

@@

\par The second method for net-vis uses the adjacency matrix of the network. The adjaceny matrix of a network, $\mathbf{A}$ is defined as:
\begin{equation}\label{eq:adjmat}
  A_{ij} =
  \begin{cases}
             1 &  \quad \text{if an edge exists } i\to j, \quad i \neq j \\
             0 & \quad \text{otherwise}
  \end{cases}
\end{equation}
%In this paper, we only consider the presence or absence of an edge, but if the network has weighted edges, $A_{ij}$ is the weight of the edge from $i \to j$. 
The adjacency matrix visualization for our toy example is also shown in Figure~\ref{fig:toyplots}. This visualization shows the network as a grid, with each row representing the ``from" or ``ego" node and each column representing the ``to" or ``alter" node of an edge. The grid space for edge $i\rightarrow j$ is filled in if $A_{ij} = 1$ and is empty if $A_{ij} = 0$. For undirected networks $\mathbf{A}$ is symmetric, so the adjacency matrix visualization is also symmetric, and each edge is drawn twice.

\par Each net-vis method has its own advantages and disadvantages. For instance, paths between two nodes in a network are easier to determine with node-link diagrams than with adjacency matrix visualizations \citep{adjmatviz}. Furthermore, in node-link diagrams, node information can be incorporated into the visualization by coloring or changing the shape of the points, and edge-level information can be incorporated by coloring the lines, or changing their thickness. Visualizing a node variable in an adjacency matrix visualization is not as simple, because this type of net-vis features the edges. An adjacency matrix visualization can, however, be useful when the network is very complex, dense, or large \citep{adjmatviz}. % Experimental studies have shown adjacency matrix visualization to be superior to node-link diagrams for large networks. For example, for basic perceptual tasks on networks, including node and edge count, adjacency matrix visualizations outperform node-link diagrams as the size and density of the network increases \citep{adjmatviz}. One drawback of the adjacency matrix visualization that \citeauthor{adjmatviz} found was that edges are overrepresented for undirected graphs, due to the symmetry of $\mathbf{A}$: the edge $i \rightarrow j$ for $i \neq j$ appears in $\mathbf{A}$ twice: in $A_{ij}$ and $A_{ji}$, and so it also appears twice in the adjacency matrix visualization. This is an advantage, however, for \emph{directed} graphs, where exactly the correct number of edges is represented in a matrix visualization. A node-link diagram, however, may underrepresent the edge count if the edges $x_{ij}$ and $x_{ji}$ both exist and are drawn on top of one another. 
Ultimately, there is no one ``correct" way to visualize network information, and we will be using both the node-link and adjacency matrix net-vis methods throughout this paper to visualize our example data and CTMC models.

@@

The continuous-time Markov chain model family is for analyzing \textit{dynamic} networks, so we can use them to study the change of a network in time while also examining the structural and node covariate efects. To define CTMCs for our purposes, we follow the general hierarchical definition outlined in \citet{saomreview} and elsewhere (see e.g.\ \citet{snijders01} and \citet{snijdersetal2010}). This hierarchy reflects our intution about social networks: most social networks, even holding constant the set of actors over time, are ever-changing as relationships form or disappear. Actors in social networks also have properties that could change their role within the network, and existing ties may affect formation or dissolution of other ties. In this section, we define the data structure, the hierarchal model, and intensity matrix of the CTMC.

@@

A \textit{dynamic network} is consists of a fixed set of $n$ nodes, that is changing over time, and is observed at $M$ discrete time points, $t_1, \dots, t_M$ with $t_1 < t_2 < \dots < t_M$. We denote the network observation at timepoint $t_m$ by $x(t_m)$ for $m = 1, \dots M$. The model assumes that this longitudinal network of discrete observations is embedded within a CTMC, which we denote by $X(T)$. The CTMC is almost entirely unobserved: we assume the first and last network observations are the starting and ending points of the CTMC, i.e. $x(t_1) \equiv X(0)$ and $x(t_M) \equiv X(\infty)$. Nearly all other steps in the chain are unobserved, with the exception of $x(t_2), \dots, x(t_{M-1})$. Unlike the first and last observations of the network, these observations do not have direct correspondence with steps in the CTMC. Thus, the observations $x(t_2), \dots, x(t_{M-1})$ are considered to be ``snapshots" of the network at some point between two steps in the CTMC.

@@

\noindent \textbf{Rate Function:} In general, the \emph{rate function}, $\rho(x, \mathbf{z}, \boldsymbol{\alpha})$, is a function of the network at its current state $x$, covariates of interest, $\mathbf{z}$, and some parameters, $\boldsymbol{\alpha}$. With this general formulation, each node can have a different rate of change. Throughout this paper, we assume a simple rate function, $\rho(x, \mathbf{z}, \boldsymbol{\alpha}) \equiv \alpha_m$ that is constant across all actors between time $t_m$ and $t_{m+1}$. With this simple rate function, we have $M-1$ rate parameters, one for each transition from $x(t_{m})$ to $x(t_{m+1})$ for $m = 1, \dots, M-1$. The rate parameter, $\alpha_m$ dictates how often an actor $i$ ``gets an opportunity" to change one of its ties, $x_{ij}$, for $j \neq i \in \{1, \dots, n\}$ in the time period from $t_{m}$ to $t_{m+1}$ for $m = 1, \dots, M-1$. The CTMC model assumes that the actors $i$ are conditionally independent given their ties, $x_{i1}, \dots, x_{in}$ at the current network state. Let $\tau(i|x,m) $ be the wait time until actor $i$ gets to the opportunity to change its ties. For any time point $T$, where $t_{m} \leq T < t_{m+1}$, the waiting time to actor $i$'s next change follows an exponential distribution,
\begin{equation} \label{eq:c1}
 \tau(i|x,m) | x_{i1}(m), \dots, x_{in}(m) \stackrel{\text{iid}}{\sim} Exp(\alpha_m)
\end{equation}
\noindent with expected value $\alpha_m^{-1}$. From Equation~\ref{eq:c1}, we can derive the waiting time to the next change opportunity by \textit{any} actor in the network, $\tau(x)$. The value $\tau(x)$ is also exponentially distributed, and has expected value $(n\alpha_m)^{-1}$. The estimation of $\alpha_m$ in \texttt{RSiena} is simple: method of moments (MoM) is used with the statistic 
\begin{equation}\label{eq:cdiff}
C_m = \sum_i\sum_j |x_{ij}(t_{m+1}) - x_{ij}(t_m)|
\end{equation}
which is the total number of changes from $x(t_m)$ to $x(t_{m+1})$.

@@

The parameters of $f_i(\boldsymbol{\beta}, x, \mathbf{Z})$ are either structural or covariate parameters. The structural parameters correspond to statistics that are \emph{only} functions of the current network state, $x$, while the covariate parameters are also functions of the covariates, $\mathbf{Z}$. According to \citet[p. 371]{snijders01}, there should be at least two parameters in the objective function: $\beta_1$ for the outdegree of nodes, and $\beta_2$ for the reciprocity of nodes. The former measures the propensity of nodes with a lot of outgoing ties to form more outgoing ties (the ``rich get richer" effect), while the latter measures the tendency of outgoing ties to be returned within a network. The statistics corresponding to these two parameters, $s_{i1}(x)$ and $s_{i2}(x)$ are given in Table~\ref{tab:effects}. In the \texttt{RSiena} software, there are over 80 possible $\boldsymbol{\beta}$ parameters with corresponding statistics $s_{ik}(x, \mathbf{Z})$ to add to the model. The definition of the statistics for all possible parameters are provided in \citet{RSienamanual}, and two additional parameters and their corresponding statistics are discussed in Section~\ref{sec:exdatamods} and shown in Figure~\ref{fig:structures}.

@@

In the CTMC, when actor $i$ can make a change, it chooses which tie $x_{i1}, \dots, x_{in}$ to change at random according to the probability $p_{ij}(\boldsymbol{\beta}, x, \mathbf{Z})$, defined via the objective function. Let $x(i \leadsto j)$ be the network identical to network $x$ with the exception of tie $x_{ij}$, which is equal to $1-x_{ij}$; i.e.\ the presence or absence of node $x_{ij}$ is flipped in the new network. The probability that the tie $x_{ij}$ changes is defined as:
\begin{equation}\label{eq:pij}
p_{ij}(\boldsymbol{\beta}, x, \mathbf{Z}) = \dfrac{\exp\left\{f_i(\boldsymbol{\beta}, x(i\leadsto j), \mathbf{Z})\right\}}{\sum_{h \neq i} \exp\left\{f_i(\boldsymbol{\beta}, x(i \leadsto h), \mathbf{Z})\right\}}
\end{equation}%When $i = j$, the numerator represents the exponential of the value of the objective function when evaluated at the current network state. 
%When the value of the objective function is high at the current state, the probability of not making a change is also high.
We explore these probabilities further in Section~\ref{sec:algorithms}.

\par \noindent \textbf{Intensity Matrix:} The probabilities $p_{ij}$ combined with the rate parameters $\alpha_m$ fully characterize the underlying CTMC that models network change. In the CTMC literature, see for instance \citet{ctmc}, chains are characterized by their  \textit{intensity matrix} $\mathbf{Q}$. This matrix describes the rate of change between two states of the CTMC process, and the rows of this matrix always add to zero. The state space for a network, denoted $\mathcal{X}$, contains $2^{n(n-1)}$ states: two possible states for an edge, $\{0,1\}$, and $n(n-1)$ edge relationships, as the network is directed and we exclude self-ties.  The intensity matrix is then a square matrix of dimension $2^{n(n-1)}$. Only one tie changes at a time in the CTMC, resulting in $n(n-1)$ reachable states from the current network state. Thus, the intensity matrix $\mathbf{Q}$ is very sparse, with only $n(n-1) + 1$ non-zero entries in each row. Note that $n(n-1)$ of these entries represent the possible states that are one edge different from a given state, while the additional non-zero entry is the diagonal entry for the current state.% All other entries in a row are structural zeroes because those network states cannot be reached from the current state in a single change.

@@

To fit a CTMC to dynamic network data, we apply the \texttt{RSiena} software, which uses simulation methods to estimate parameter values using either MoM or maximum likelihood (ML) estimation \citep{RSiena}. We chose to use MoM estimation, so we describe that process here, while more information on ML estimation can be found in \citet{SienaAlgs}. %because the theory behind it was established in \citet{saompaper}, while the maximum likelihood estimation methods were not fully established until \citet{saomsml2010}, though \texttt{RSiena} contains capabilities to use maximum likelihood estimation. We also use the score function method for estimating the derivatives of the expected values, as opposed to the finite differences method, both of which are outlined in detail in .
The underlying SIENA software uses a Robbins-Monro algorithm (see \citet{robbinsmonro}) to estimate the solution of the moment equation
\begin{equation}\label{eq:mom}
E_{\theta}S = s_{obs}
\end{equation}
where $\theta = (\boldsymbol{\alpha}, \boldsymbol{\beta})$ is the vector of rate and objective function parameters, and $s_{obs}$ is the observed vector of the corresponding statistics summed over all network observations. Estimation conditions on the first observed network, $x(t_1)$, then proceeds to estimate model parameters for all future observations. % awkward. fix it. 
The full SIENA MoM estimation algorithm operates in three phases, as described in \citet{RSienamanual} and \citet{SienaAlgs}, and we briefly summarize these phases here. The first phase obtains initial estimates of the score functions for use in the second phase. The second phase is the Robbins-Monro algorithm, and results in parameter estimates through iterative updates and simulation from the CTMC at the iterations of the parameter values. The third phase uses the parameters from phase two to estimate the score functions and covariance matrix of the parameter estimates and carries out convergence checks. In Section~\ref{sec:algorithms}, we further explore these phases in the SIENA MoM algorithm through visualization. %, bringing them out of their ``black box" and into the light.

>>>>

Dynamic network models do not have as rich of a history as static network models, largely due to the historical lack of data and available analytical tools \citep{goldenberg09}. With the advent of social media and ever-increasing computational power, however, more and more researchers are turning their attention toward dynamic network models.  Some of the classical models, such as the Erd\"{o}s-R\'{e}nyi model and the small-world model of \citet{watts} are pseudo-dynamic because they could be applied in a dynamic way but rarely are \citep{goldenberg09}.  
% talk up dynamic social network models. they are much more realistic. they don't treat nodes as exchangeable, they allow the nodes to affect the ties. They account for both structural and covariate effects. They are very flexible. They are the ideal model because they are the only ones that have any chance of actually capturing the mechanisms. 
Fully dynamic social network models, however, are more complicated: they incorporate both the network structure and its changes in time.  If we can capture this change mechanism in time, we can truly begin to model the underlying network dynamics that drive social connections. A comprehensive review of dynamic network models is presented in \citet{goldenberg09}.

\par We are particularly interested one truly dynamic family of network models: continuous time Markov chain (CTMC) models, first introduced in \citet{ctmcorig}. We chose these models because they correspond to our intuition about how social networks are formed, and they are very flexible and widely applicable. In these models, network changes are modeled hierarchically at the node level: nodes are given opportunities to change via a rate function, then the node chooses which edge to change according to its utility function. %first step is the changes happen at the smallest possible level (one edge between two nodes), and subsequent changes condition on past changes. 
In each step of the hierarchy, many possible effects can be implemented to capture the change mechanism. For example,  the rate function could be a simple waiting-time model, or it could depend on a node's indegree or outdegree: we may have reason to believe that the more popular nodes change their ties more frequently. In addition, we can incorporate node covariate information into the hierarchy, such as gender or education level of people in a social network. For instance, the genders of two people could effect whether and how quickly they establish a relationship, and people will likely form more relationships with those who share their education level. Thus, the transition probabilities between two states of a network depend on the network's structure, characteristics of the nodes, or both \citep{saompaper}.  For example, \citet{depression} use a CTMC model to learn about the tendency of depressed adolescents to form friendships and found that depressed students tended to avoid creating new friendships over time. The CTMC model family can also be expanded to model node behavior, as in \citet{chyzh2016}, who applies CTMC models to show that more indirect trade relationships in a country decrease its overall respect for human rights. % This means that node-level covariate effects, in addition to commonly used effects such as reciprocity or transitivity, can be incorporated into the model. 
% Cite some of olga's work. what can you do with these models? cite other work with social networks. 
Thus, with CTMC models we can examine assumptions made about social networks. %For example, we may assume that teenagers who drink heavily are more likely to be friends with other teenagers who drink heavily.  We can assess our assumptions on shared interests into the modeling process through a particular subfamily of CTMC models known as stochastic actor-oriented models (SAOMs) \citep{saompaper}.

\par With the flexibility of CTMC models comes increased complexity and estimation difficulty \citep{saomreview}. Likelihood functions quickly become very complex due to the inherent dependency structure in the data, and therefore computational methods, such as Markov chain Monte Carlo (MCMC), are used to fit models. Methods for maximum likelihood estimation (MLE) for these models were not introduced until \citet{saomsml2010}, and method of moments (MM) estimation is frequently used \citep{RSiena}. %SAOMs are typically fit to dynamic social network data using a series of MCMC phases for finding method of moments (MoM) estimates of model parameters. 
For CTMC parameter estimation, we use the software SIENA, and its \texttt{R} implementation \texttt{RSiena} \citep{rstats, RSiena}. This software is a considerable contribution to the field of social network analysis, allowing for estimation of structural effects, node covariate effects, as well as modeling both network and node changes in a wide variety of CTMC models. This extensive parameter estimation process is largely hidden from the user, and the primary outputs are parameter estimates. Direct simulation from a given network model with fixed parameter values is also made easy with \texttt{RSiena}. This software opens up a great deal of possibilities for network modeling and simulation.% need to emphasize the need for visualization.

% ok. we know why ctmcs but WHY visualize them? where is the motivation?? who cares and why? 
% talk about lack of goodness of fit diagnostics. 
Since we chose the CTMC models for dynamic social networks, we would like to determine how well they fit data. We chose these models because they match our intuition about how social networks form in reality, and after we fit a model, we will want to assess its goodness-of-fit. In statistical network models generally, the lack of large-sample theory leads to a lack of goodness-of-fit tests \citep{goldenberg09}. Most methods for assessing goodness-of-fit for network models, therefore, involve comparing statistics computed on simulated networks to the same statistics computed on the data, for example in \citet{hunter2008}. More recently, \citet{gofschwein} used score-type tests for forward model selection in the CTMC model context. The \texttt{RSiena} software also contains methods for goodness-of-fit based on network statistics, such as outdegree distribution, where the data values are compared to simulated values as in \citet{hunter2008} \citep{RSiena}. % need to transition to model vis. % say something about how important visualization is in network analysis.

We propose new methods for goodness-of-fit and other diagnostics that use visualization as a supplement to computational methods. Network visualization, also called graph drawing, is a prominent subfield of network analysis with a rich history: early drawings of complete graphs date back to 1280 CE \citep{knuth2000}. We use this visualization foundation to combine modeling with visualization. Note that we do not propose new methods of visualizing networks, and for a complete review of the field of graph drawing we direct the reader to \citet{graphdrawing}. What is novel in our work is the use of common visualization methods to diagnose network model with model visualization

\emph{Model visualization} complements traditional model diagnostic tools such as $R^2$ values and residual plots to enrich our understanding and interpretation of statistical models \citep{modelvis}. We apply the three principles of model-vis: view the model in the data space, visualize collections of models, and explore the process of fitting the model, to the models and data introduced in Section~\ref{sec:exdatamods}.  By translating the model visualization work of \citet{modelvis} to the dynamic network model space, we provide network researchers with a new suite of tools for visualizing and diagnosing performance of their network models. Though we apply the tools to CTMCs because we find the models intuitive, the methods we use can be applied to any network model, static or dynamic.% do something awesome

@@

%\noindent \textbf{Network Data Structure:} 
Network data across fields have similar structure: they consist of units of observation and connections between those units. In a social network, the units of observations, called \textit{nodes} or \textit{actors}, are people, while the connections, called \textit{edges} or \textit{ties}, are their relationships with each other. %Networks change over time, like when new relationships are formed in a social network. 
Other network data include inherent variables of interest on the nodes, such as age or gender, and the type of relationship (friends, coworkers, etc) that an edge represents. Dynamic networks are observed at many discrete points in time, where each observation of a network is often called a ``wave" of data.
%These two network data structures pose unique problems to analysts, such as, ``How does the strength of a tie between two nodes affect the overall structure of the network?" or ``Do differences in node variables affect the formation of edges?" We propose to answer these questions and more through visual exploration of network data and models. \\

@@

\par The second method for net-vis uses the adjacency matrix of the network. The adjaceny matrix of a network, $\mathbf{A}(x)$ is defined as:
\begin{equation}\label{eq:adjmat}
  A_{ij}(x) =
  \begin{cases}
             1 &  \quad \text{if } x_{ij} = 1, \quad i \neq j \\
             0 & \quad \text{otherwise}
  \end{cases}
\end{equation}
%In this paper, we only consider the presence or absence of an edge, but if the network has weighted edges, $A_{ij}$ is the weight of the edge from $i \to j$. 
The adjacency matrix visualization for our toy example is also shown in Figure~\ref{fig:toyplots}. This visualization shows the network as a grid, with each row representing the ``from" or ``ego" node and each column representing the ``to" or ``alter" node of an edge. The grid space for $x_{ij}$ (the edge $i\rightarrow j$) is filled in if $A_{ij} = 1$ and is empty if $A_{ij} = 0$. For undirected networks $\mathbf{A}$ is symmetric, so the adjacency matrix visualization is also symmetric, with each edge drawn twice.

\par Each net-vis method has its own advantages and disadvantages. For instance, paths between two nodes in a network are easier to determine with node-link diagrams than with adjacency matrix visualizations \citep{adjmatviz}. Furthermore, in node-link diagrams, node information can be incorporated into the visualization by coloring or changing the shape of the points, and edge-level information can be incorporated by coloring the lines or changing their thickness. Visualizing a node variable in an adjacency matrix visualization is not as simple, because this type of net-vis features the edges. An adjacency matrix visualization can, however, be useful when the network is very complex, dense, or large \citep{adjmatviz}. % Experimental studies have shown adjacency matrix visualization to be superior to node-link diagrams for large networks. For example, for basic perceptual tasks on networks, including node and edge count, adjacency matrix visualizations outperform node-link diagrams as the size and density of the network increases \citep{adjmatviz}. One drawback of the adjacency matrix visualization that \citeauthor{adjmatviz} found was that edges are overrepresented for undirected graphs, due to the symmetry of $\mathbf{A}$: the edge $i \rightarrow j$ for $i \neq j$ appears in $\mathbf{A}$ twice: in $A_{ij}$ and $A_{ji}$, and so it also appears twice in the adjacency matrix visualization. This is an advantage, however, for \emph{directed} graphs, where exactly the correct number of edges is represented in a matrix visualization. A node-link diagram, however, may underrepresent the edge count if the edges $x_{ij}$ and $x_{ji}$ both exist and are drawn on top of one another. 
Ultimately, there is no one ``correct" way to visualize network information, and we will be using both the node-link and adjacency matrix net-vis methods throughout this paper to visualize our example data and CTMC models.

@@

The continuous-time Markov chain (CTMC) model family is for analyzing dynamic networks, so we can use them to study the change of a network in time while also examining the structural and node covariate efects. To define CTMCs for our purposes, we follow the general hierarchical definition outlined in \citet{saomreview} and elsewhere (see e.g.\ \citet{snijders01} and \citet{snijdersetal2010}). This hierarchy reflects our intution about social networks: most social networks, even holding constant the set of actors over time, are ever-changing as relationships form or disappear. Actors in social networks also have properties that could change their role within the network, and existing ties may affect formation or dissolution of other ties. In this section, we define the data structure, the hierarchal model, and intensity matrix of the CTMC.

@@

A dynamic network consists of a fixed set of $n$ nodes that is changing over time, and is observed at $M$ discrete time points, $t_1, \dots, t_M$ with $t_1 < t_2 < \dots < t_M$. We denote the network observation at timepoint $t_m$ by $x(t_m)$ for $m = 1, \dots M$. The model assumes that this longitudinal network of discrete observations is embedded within a CTMC, which we denote by $X(T)$. The CTMC is almost entirely unobserved: we assume the first and last network observations are the starting and ending points of the CTMC, i.e. $x(t_1) \equiv X(0)$ and $x(t_M) \equiv X(\infty)$. Nearly all other steps in the chain are unobserved, with the exception of $x(t_2), \dots, x(t_{M-1})$. Unlike the first and last observations of the network, $x(t_2), \dots, x(t_{M-1})$ do not have direct correspondence with steps in the CTMC. Thus, the observations $x(t_2), \dots, x(t_{M-1})$ treated as ``snapshots" of the network at some point between two steps in the CTMC.

@@

\noindent \textbf{Rate Function:} In general, the \emph{rate function}, $\rho(x, \mathbf{z}, \boldsymbol{\alpha})$, is a function of the network at its current state $x$, covariates of interest, $\mathbf{z}$, and some parameters, $\boldsymbol{\alpha}$. With this general formulation, each node can have a different rate of change. Throughout this paper, we assume a simple rate function, $\rho(x, \mathbf{z}, \boldsymbol{\alpha}) \equiv \alpha_m$ that is constant across all actors between time $t_m$ and $t_{m+1}$. With this simple rate function, we have $M-1$ rate parameters, one for each transition from $x(t_{m})$ to $x(t_{m+1})$ for $m = 1, \dots, M-1$. The rate parameter, $\alpha_m$ dictates how often an actor $i$ is selected to change one of its ties, $x_{ij}$, for $j \neq i \in \{1, \dots, n\}$ in the time period from $t_{m}$ to $t_{m+1}$ for $m = 1, \dots, M-1$. The CTMC model assumes that the actors $i$ are conditionally independent given their ties $x_{i1}, \dots, x_{in}$ at the current network state. Let $\tau(i|x,m)$ be the wait time until actor $i$ gets to the opportunity to change its ties. For any time point $T$, where $t_{m} \leq T < t_{m+1}$, the waiting time to actor $i$'s next change follows an exponential distribution,
\begin{equation} \label{eq:c1}
 \tau(i | x_{i1}(m), \dots, x_{in}(m)) \stackrel{\text{iid}}{\sim} Exp(\alpha_m)
\end{equation}
\noindent with expected value $\alpha_m^{-1}$. From Equation~\ref{eq:c1}, we can derive the waiting time to the next change opportunity by \textit{any} actor in the network, $\tau(x)$. The value $\tau(x)$ is also exponentially distributed, and has expected value $(n\alpha_m)^{-1}$. The estimation of $\alpha_m$ in \texttt{RSiena} is simple: method of moments (MoM) is used with the statistic 
\begin{equation}\label{eq:cdiff}
C_m = \sum_i\sum_j |x_{ij}(t_{m+1}) - x_{ij}(t_m)|,
\end{equation}
which is the total number of edge differences between $x(t_m)$ and $x(t_{m+1})$.

@@

The parameters of $f_i(\boldsymbol{\beta}, x, \mathbf{Z})$ are either structure or covariate parameters. The structure parameters correspond to statistics that are \emph{only} functions of the current network state, $x$, while the covariate parameters are also functions of the covariates, $\mathbf{Z}$. According to \citet[p. 371]{snijders01}, there should be at least two parameters in the objective function: $\beta_1$ for the outdegree of nodes, and $\beta_2$ for the reciprocity of nodes. The former measures the propensity of nodes with a lot of outgoing ties to form more outgoing ties (the ``rich get richer" effect), while the latter measures the tendency of outgoing ties to be returned within a network. The statistics corresponding to these two parameters, $s_{i1}(x)$ and $s_{i2}(x)$ are given in Figure~\ref{fig:structures} with visualizations of the corresponding effects. In the \texttt{RSiena} software, there are over 80 possible $\boldsymbol{\beta}$ parameters with corresponding statistics $s_{ik}(x, \mathbf{Z})$ to add to the model. The definition of the statistics for all possible parameters are provided in \citet{RSienamanual}, and two additional parameters and their corresponding statistics are discussed in Section~\ref{sec:exdatamods} and shown in Figure~\ref{fig:structures}.

@@

In the CTMC, when actor $i$ can make a change, it chooses which tie $x_{i1}, \dots, x_{in}$ to change at random according to the probability $p_{ij}(\boldsymbol{\beta}, x, \mathbf{Z})$. Let $x(i \leadsto j)$ be the network identical to network $x$ with the exception of tie $x_{ij}$, which is equal to $1-x_{ij}$; i.e.\ the presence or absence of node $x_{ij}$ is flipped in the new network. The probability that the tie $x_{ij}$ changes is defined as:
\begin{equation}\label{eq:pij}
p_{ij}(\boldsymbol{\beta}, x, \mathbf{Z}) = \dfrac{\exp\left\{f_i(\boldsymbol{\beta}, x(i\leadsto j), \mathbf{Z})\right\}}{\sum_{h \neq i} \exp\left\{f_i(\boldsymbol{\beta}, x(i \leadsto h), \mathbf{Z})\right\}}
\end{equation}%When $i = j$, the numerator represents the exponential of the value of the objective function when evaluated at the current network state. 
%When the value of the objective function is high at the current state, the probability of not making a change is also high.
We explore these probabilities further in Section~\ref{sec:algorithms}.

\par \noindent \textbf{Intensity Matrix:} The probabilities $p_{ij}$ combined with the rate parameters $\alpha_m$ fully characterize the underlying CTMC that models network change. In the CTMC literature, see for instance \citet{ctmc}, chains are characterized by their  \textit{intensity matrix} $\mathbf{Q}$. This matrix describes the rate of change between two states of the CTMC process, and the rows of this matrix always add to zero. The state space for a network, denoted $\mathcal{X}$, contains $2^{n(n-1)}$ states: there are two possible states for an edge, $\{0,1\}$, and $n(n-1)$ possible edge relationships, as the network is directed and we exclude self-ties.  The intensity matrix is then a square matrix of dimension $2^{n(n-1)}$. Only one tie changes at a time in the CTMC, resulting in $n(n-1)$ reachable states from the current network state. Thus, the intensity matrix $\mathbf{Q}$ is very sparse, with only $n(n-1) + 1$ non-zero entries in each row. Note that $n(n-1)$ of these entries represent the possible states that are one edge different from a given state, while the additional non-zero entry is the diagonal entry for the current state.% All other entries in a row are structural zeroes because those network states cannot be reached from the current state in a single change.

@@

To fit a CTMC to dynamic network data, we apply the \texttt{RSiena} software, which uses simulation methods to estimate parameter values using either MoM or maximum likelihood (ML) estimation \citep{RSiena}. We chose to use MoM estimation, so we describe that process here, while more information on ML estimation can be found in \citet{SienaAlgs}. %because the theory behind it was established in \citet{saompaper}, while the maximum likelihood estimation methods were not fully established until \citet{saomsml2010}, though \texttt{RSiena} contains capabilities to use maximum likelihood estimation. We also use the score function method for estimating the derivatives of the expected values, as opposed to the finite differences method, both of which are outlined in detail in .
The underlying SIENA software uses a Robbins-Monro algorithm (see \citet{robbinsmonro}) to estimate the solution of the moment equation
\begin{equation}\label{eq:mom}
E_{\theta}S = s_{obs}
\end{equation}
where $\theta = (\boldsymbol{\alpha}, \boldsymbol{\beta})$ is the vector of rate and objective function parameters, and $s_{obs}$ is the observed vector of the corresponding statistics summed over all network observations. Estimation conditions on the first observed network, $x(t_1)$, then proceeds to estimate model parameters for all future observations. % awkward. fix it. 
The full SIENA MoM estimation algorithm operates in three phases, as described in \citet{RSienamanual} and \citet{SienaAlgs}, which we briefly summarize here. The first phase obtains initial estimates of the score functions for use in the second phase. The second phase is the Robbins-Monro algorithm, and results in parameter estimates through iterative updates and simulation from the CTMC at the iterations of the parameter values. The third phase uses the parameters from phase two to estimate the score functions and covariance matrix of the parameter estimates and carries out convergence checks. In Section~\ref{sec:algorithms}, we further explore these phases in the SIENA MoM algorithm through visualization. For a more detailed description of the fitting process, we direct the reader to \citet{RSienamanual, SienaAlgs}%, bringing them out of their ``black box" and into the light.
